\documentclass[11pt]{article}
\usepackage[cm]{fullpage}
%%AVC PACKAGES
\usepackage{avcgreek}
\usepackage{avcfonts}
\usepackage{avcmath}
\usepackage[numberby,skip=9pt plus 2pt minus 7pt]{avcthm}
\usepackage{qcmacros}
\usepackage{goldstone}
%%MACROS FOR THIS DOCUMENT
\usepackage{accents}
\newcommand{\oc}[1]{\ensuremath{\accentset{\circ}{#1}}}

\begin{document}

\section*{OMP2}


\begin{enumerate}
\item
Build $h_p^q$ and $\ol{g}_{pq}^{rs}$ in the Hartree-Fock spin-orbital basis.

\item
Build an empty array of amplitudes, $t_{ab}^{ij}=0$.

\item
\label{item:recursive-step}
Build the Fock matrix.
\begin{align}
  f_p^q
=
  h_p^q
+
  \ol{g}_{pi}^{qi}
\end{align}

\item
Build the off-diagonal Fock matrix and the orbital energies.
\begin{align}
  \ev_p
=
  f_p^p
&&
  f'{}_p^q
=
  (1 - \d_p^q)
  f_p^q
\end{align}

\item
\label{itemocc-step-one}
Update the amplitudes.
\begin{align}
  t_{ab}^{ij}
=
  (\mc{E}_{ab}^{ij})^{-1}
  \pr{
    \ol{g}_{ab}^{ij}
  +
    P_{(a/b)}
    f'{}_a^c
    t_{cb}^{ij}
  -
    P^{(i/j)}
    f'{}_k^i
    t_{ab}^{kj}
  }
\end{align}

\item
Build the one- and two-particle density matrices.
\begin{align}
  \g^p_q
=
  \tl{\g}^p_q
+
  \oc{\g}^p_q
&&
  \g^{pq}_{rs}
=
  \tl{\g}^{pq}_{rs}
+
  P^{(p/q)}_{(r/s)}
  \tl{\g}^p_r\,
  \oc{\g}^q_s
+
  P_{(r/s)}
  \oc{\g}^p_r
  \oc{\g}^q_s
\end{align}
\begin{align}
  \tl{\g}^a_b
=
  \tfr{1}{2}
  t^{ac*}_{ij}
  t_{bc}^{ij}
&&
  \tl{\g}^i_j
=
-
  \tfr{1}{2}
  t^{ab*}_{jk}
  t_{ab}^{ik}
&&
  \tl{\g}_{ij}^{ab}
=
  t_{ij}^{ab*}
&&
  \tl{\g}_{ab}^{ij}
=
  t_{ab}^{ij}
&&
  \oc{\g}^p_q
=
\left\{
\begin{array}{ll}
  \d^i_j
&
  \text{for $p=i$, $q=j$}
\\
  0
&
  \text{otherwise}
\end{array}
\right.
\end{align}

\item
Compute the Newton-Raphson step.
\begin{align}
  x_a^i
=
  \fr{(\bo{F} - \bo{F}\dg)_a^i}{\ev_i - \ev_a}
&&
  (\bo{F})_p^q
\equiv
  h_p^r
  \g^q_r
+
  \tfr{1}{2}
  \ol{g}_{pr}^{st}
  \g^{qr}_{st}
\end{align}

\item
Build the Newton-Raphson orbital rotation matrix.
\begin{align}
  \bo{U}
=
  \mr{exp}(\bo{X} - \bo{X}\dg)
&&
  \bo{X}_\mr{vo}
=
  [x_a^i]
\end{align}

\item
Rotate the spin-orbital coefficients.
\begin{align}
  \bo{C}
\leftarrow
  \bo{C}\bo{U}
\end{align}

\item
Transform the one- and two-electon integrals to the spin-orbital basis using the new coefficient matrix.
\begin{align}
  h_p^q
=
  \sum_{\mu\nu}
  C_{\mu p}^*\,
  h_{\mu\nu}\,
  C_{\nu q}
&&
  \ol{g}_{pq}^{rs}
=
  \sum_{\mu\nu\rh\si}
  C_{\mu p}^*\,
  C_{\nu q}^*\,
  \ip{\mu \nu||\rh\si}
  C_{\rh r}
  C_{\si s}
\end{align}

\item
Evaluate the energy.
\begin{align}
  E
=
  h_p^q
  \g^p_q
+
  \tfr{1}{4}
  \ol{g}_{pq}^{rs}
  \g^{pq}_{rs}
\end{align}

\item
If the energy is converged, quit.
Otherwise, return to step~\ref{item:recursive-step}.
\end{enumerate}


\end{document}