\documentclass[11pt]{article}
\usepackage[cm]{fullpage}
%%AVC PACKAGES
\usepackage{avcgreek}
\usepackage{avcfonts}
\usepackage{avcmath}
\usepackage[numberby=section,skip=9pt plus 2pt minus 7pt]{avcthm}
\usepackage{qcmacros}
\usepackage{goldstone}
%%MACROS FOR THIS DOCUMENT
\numberwithin{equation}{section}
\usepackage[
  margin=1.5cm,
  includefoot,
  footskip=30pt,
  headsep=0.2cm,headheight=1.3cm
]{geometry}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE,RO]{Quiz 7, Handout 1: Perturbative analysis}
\fancyfoot[CE,CO]{\thepage}
\usepackage{url}
\makeatother
\newcommand{\resolventline}[2][1]{
  \tikz[overlay]{
      \draw[thick,flexdotted] (0,-1ex) to ++(0,#1*4.5ex) node[above,inner sep=1pt] {#2};
  }
}

\begin{document}


\appendix
\section{Fa\`a di Bruno's formula}

\begin{thm}
\thmtitle{Fa\`a di Bruno's formula}
\begin{align}
  \pd{^n}{x_1\cd \pt x_n}
  f(g(\bm{x}))
=
  \sum_{k=1}^n
  \sum_{(\bm{x}_1,\ld,\bm{x}_k)}^{\mc{P}_k(\bm{x})}
  f\ord{k}(g(\bm{x}))
  \prod_{i=1}^k
  \pd{
    ^{|\bm{x}_i|}
    g(\bm{x})
  }{
    x_{i,1}
  \cd
    \pt
    x_{i,|\bm{x}_i|}
  }
\end{align}
\end{thm}



\section{Frantz-Mills factorization theorem}

\begin{dfn}
\label{dfn:resolvent-graph}
\thmtitle{Resolvent graph}
A \textit{resolvent graph} represents a product of operators and resolvents.
Formally, we will restrict this term to graphs in which each resolvent line spans the width of the diagram.
Such diagrams can be broken into \textit{levels}, which are the vertical spaces between resolvent lines, numbered from bottom to top with zero indexing.
Consequently, an operator lies in the $k\eth$ level if there are $k$ resolvent lines below it.
Formally, then, an \textit{$m$-level resolvent graph} $G(\rh,m)\equiv(G,\rh,m)$ associates each operator $o$ in $G$ with a specific level $\rh(o)=\rh_o$ in $\mb{Z}_m=\{0,1,\ld,m-1\}$ through the \textit{level map} $\rh$.\,\footnote{
  Note that an $m$-level resolvent graph contains $m-1$ resolvents.
}
Each line $l$ in $G$ crosses resolvents
$
  \mr{min}(\rh_{h(l)},\rh_{t(l)}) + 1
$
through
$
  \mr{max}(\rh_{h(l)},\rh_{t(l)})
$.
\end{dfn}

\begin{rmk}
Many graphs containing resolvent lines will not meet our formal definition in \cref{dfn:resolvent-graph}, but note that we can always collapse the subgraphs that do into intermediate operators.
In this way, an arbitrary graph with resolvent lines can be recursively defined as a \textit{resolvent graph of resolvent graphs}.
\end{rmk}

\begin{rmk}
A product of graphs $G=(L,O,h,t)$ and $G'=(L',O',h',t')$ is itself a graph, which is formally given by
\begin{align}
  GG'
=
  (L\cup L', O\cup O', h\oplus h', t\oplus t')
\end{align}
where $h\oplus h'$ acts as $h$ on lines from $L$ and as $h'$ on lines from $L'$.
The combined tail function is defined similarly.
\end{rmk}

\begin{ntt}
Given
$
  G(\rh,m)
$,
let
$
  G_k(\rh_k,k+1)
$
denote that we have collapsed the subgraph formed from $k$ and up into an intermediate operator
$
  G[O_{\geq k}](\rh-k,m-k)
$
into an intermediate operator
$
  o_{\geq k}
$
in level
$
  \rh_k(o_{\geq k})
=
  k
$
and with
$
  \rh_k(o)
=
  \rh(o)
$
for all
$
  o
\in
  O
  \setminus
  O_{\geq k}
$.
\end{ntt}

\begin{dfn}
\label{dfn:resolvent-graph}
\thmtitle{Zipper graph}
A \textit{$(k,k')$-joined zipper graph} of $G(\rh,m)$ and $G'(\rh',m')$ 
\begin{align}
  G_k
  G_{k'}'
  (
    \rh_\pi^{k,k'},\
    k+k'
  )
&&
  \rh
=
\left\{
\begin{array}{r@{\ }l}
\end{array}
\right.
&&
  \pi
\in
  \mr{S}_{\mb{Z}_{k+k'}}^{(k,k')}
\end{align}
\end{dfn}

\begin{dfn}
\label{dfn:insertion-graph}
\thmtitle{Insertion graph}
\end{dfn}

\begin{dfn}\label{dfn:resolvent-graph}
\thmtitle{Resolvent graph}
A \textit{resolvent graph} represents a normal-ordered product of operators and resolvents.
Graphs with disconnected parts that don't share any resolvent lines are considered products of separate resolvent graphs.
Vertical spaces between resolvent lines in a resolvent graph are termed \textit{levels}, which are numbered from bottom with zero indexing.
Therefore, an operator lies in the $k\eth$ level if there are $k$ resolvent lines below it.
Formally, then, an \textit{$m$-level resolvent graph} $G(\rh,m)\equiv(G,\rh,m)$ associates each operator $o$ in $G$ with a specific level $\rh(o)=\rh_o$ in $\mb{Z}_m=\{0,1,\ld,m-1\}$ through the \textit{level map} $\rh$.\,\footnote{
  Note that an $m$-level resolvent graph contains $m-1$ resolvents.
}
Therefore, each line $l$ in $G$ crosses resolvents
$
  \mr{min}(\rh_{h(l)},\rh_{t(l)}) + 1
$
through
$
  \mr{max}(\rh_{h(l)},\rh_{t(l)})
$.
\end{dfn}



\begin{dfn}
According to definition~\ref{dfn:resolvent-graph}, however, a product of resolvent graphs is not a resolvent graph.
There are several \ld\\
The \textit{combination graphs} of $G(\rh,m)$ and $G'(\rh',m')$ have the form
$GG'(\rh_{\pi,\si}^{k,k'},m+m'-1)$\\
\begin{align*}
\begin{array}{r@{\ }l}
  \rh_{\pi,\si}^{k,k'}(o)
=&
\left\{
\begin{array}{ll}
  \pi(\rh(o))      & \rh(o)<k \\
  k + k'           & \rh(o)=k \\
  \makebox[2.8cm][l]{$\si(\rh(o) + k')$} & \rh(o)>k \\
\end{array}
\right.
\\[20pt]
  \rh_{\pi,\si}^{k,k'}(o')
=&
\left\{
\begin{array}{ll}
  \pi(\rh'(o') + k)   & \rh(o')<k' \\
  k + k'              & \rh(o')=k' \\
  \makebox[2.8cm][l]{$\si(\rh'(o') + m-1)$} & \rh(o')>k' \\
\end{array}
\right.
\end{array}
\hspace{20pt}
\begin{array}{r@{\ }l}
  o\,
\in&
  O
\\[5pt]
  o'
\in&
  O'
\\[5pt]
  \pi\,
\in&
  \mr{L}_{k,k'}
\\[5pt]
  \si\,
\in&
  \mr{U}_{k,k'}^{m,m'}
\end{array}
\end{align*}
where
$
  \mr{L}_{k,k'}
\equiv
  \mr{S}^{(k,k')}_{\mb{Z}_{k+k'}}
$
and
$
  \mr{U}_{k,k'}^{m,m'}
\equiv
  \mr{S}^{(m-k-1,m'-k'-1)}_{\mb{Z}_{m'+m'-k-k'-2} + k + k' + 1}
$
are interleavings of the levels below and above level $k$ and $k'$ in the respective graphs.
\end{dfn}

\begin{thm}
\thmtitle{Frantz-Mills Factorization Theorem}
\begin{align}
  G(\rh,m)
  G'(\rh',m')
=
  \sum_\pi^{\mr{L}_{k,k'}}
  GG'(\rh_{\pi,\si}^{k,k'},m+m'-1)
&&
\begin{array}{r@{\ }l@{\,}l}
  0\leq& k &<m \\[2pt]
  0\leq& k'&<m'
\end{array}
\hspace{10pt}
  \si\in\mr{U}_{k,k'}^{m,m'}
\end{align}
\end{thm}

\begin{dfn}
\thmtitle{Insertion graph}
\end{dfn}

%\begin{dfn}
%\thmtitle{Insertion graph}
%The graphs encountered in perturbation theory contain operators separated by resolvents.
%Two disconnected parts are termed \textit{separate} if no resolvent line crosses both.
%Vertical spaces between resolvent lines are termed \textit{levels}, which we number from %top to bottom for each separate part.
%A pair of resolvents with no intervening operators encloses an \textit{empty level}.
%Inserted brackets in the bracketing expansion produce separate and unlinked \textit{insertion graphs}.
%The empty level in the \textit{remainder} created by the insertion is the \textit{level of the insertion}.
%\end{dfn}



\end{document}